---
title: "Assignment 6 - Web APIs"
author: "Ryan Gordon"
date: "03/31/2019"
output:
  html_document:
    code_folding: hide
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float:
      collapsed: false
---

## Load All Necessary Libraries

Here are a list of the necessary packages for this assignment:

* RCurl - Save the URL
* jsonlite - Parse JSON file
* DT - Create datatable

```{r load-libraries, eval=TRUE, warning=FALSE, message=FALSE}
library(RCurl)
library(jsonlite)
library(DT)
```

## Web API

For this assignment, I went ahead and retrieved some information about critics and displayed as a datatable. I am getting a weird warning every time I open in a web browser or compile. I have not been able to get rid of it.

```{r get-data, eval=TRUE, comment=NA, warning=FALSE, message=FALSE}
nyt_url <- "https://api.nytimes.com/svc/movies/v2/critics/all.json"
nyt_api <- "2clgLwkWHqtVThCH1xNtVQD906tFnVmg"
nyt_url_comp <- paste0(nyt_url,"?api-key=",nyt_api)
critics <- data.frame(fromJSON(nyt_url_comp),flatten=TRUE)
datatable(critics)
```