---
title: "Assignment 5 - Working with XML and JSON in R"
author: "Ryan Gordon"
date: "03/17/2019"
output:
  html_document:
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float:
      collapsed: false
---

## Loading Libraries

Here are a list of the necessary packages for this assignment:

* RCurl - Save the URL
* XML - Parse HTML and XML files
* jsonlite - Parse JSON file
* DT - Create datatable

```{r load-libraries, eval=TRUE, warning=FALSE, message=FALSE}
library(RCurl)
library(XML)
library(jsonlite)
library(DT)
```

## HTML, XML and JSON Web Scraping {.tabset .tabset-fade}

Each of the tabs represents a particular file type, where the data was scraped and transformed into a data frame using R. All of the information in each of these files are the same as the other two files, but are saved in their respective file types.

### HTML

For the HTML file, the readHTMLTable function was used to parse the data. This reads the file in as a "list," and this list contains a dataframe, "ChemEngBooks," as shown below:

```{r html-grab, eval=TRUE, warning=FALSE, comment=NA}
books.html.url <- getURL("https://raw.githubusercontent.com/rg563/DATA607/master/Assignments/Assignment%205/books.html")
books.html <- readHTMLTable(books.html.url,header=TRUE)
class(books.html)
class(books.html$ChemEngBooks)
books.html
```

Next, we simply extract the ChemEngBooks from "books.html" since this is the data frame.

```{r html-df, eval=TRUE, warning=FALSE, comment=NA}
books.html.df <- books.html$ChemEngBooks
books.html.df
```

Finally, we can display the data frame as a data table using the datatable function.

```{r html-datatable, eval=TRUE, warning=FALSE, comment=NA}
datatable(books.html.df)
```

### XML

For the XML file, I used xmlParse to read the data into a variable. We did not see the function return a "list" this time, but we saw it return a "XMLInternalDocument." As you can see below, it appears that the entire XML file was just reprinted.

```{r xml-grab, eval=TRUE, warning=FALSE, comment=NA}
books.xml.url <- getURL("https://raw.githubusercontent.com/rg563/DATA607/master/Assignments/Assignment%205/books.xml")
books.xml <- xmlParse(books.xml.url)
class(books.xml)
books.xml
```

We can use the xmlRoot function to remove the XML header and return the top-level XMLNode, which in this case is the ChemEngBooks itself.

```{r xml-root, eval=TRUE, warning=FALSE, comment=NA}
books.xml.root <- xmlRoot(books.xml)
books.xml.root
```

We can then utilize the xmlValue function to retrieve all the important information of each node. If we use a nested xmlSApply function we can retrieve all the information from each book into a matrix. The reason we need to use a nested one is because the ChemEngBooks are the top-element, followed by each Book node, and then the values within each Book node are the important information.

```{r xml-matrix, eval=TRUE, warning=FALSE, comment=NA}
books.xml.matrix <- xmlSApply(books.xml.root, function(x) xmlSApply(x, xmlValue))
books.xml.matrix
```

Finally, we need to transpose the matrix, and then turn it into a dataframe.

```{r xml-df, eval=TRUE, warning=FALSE, comment=NA}
books.xml.matrix.t <- t(books.xml.matrix)
books.xml.df <- data.frame(books.xml.matrix.t, row.names = NULL)
books.xml.df
```

The final data table from the XML file is shown below:

```{r xml-datatable, eval=TRUE, warning=FALSE, comment=NA}
datatable(books.xml.df)
```

### JSON

The procedure for scraping a JSON file is almost identical to the HTML file with the exception of the fromJSON function used to read the file. Everything else about the procedures and structure of R code is the same as the HTML file.

```{r json-grab, eval=TRUE, warning=FALSE, comment=NA}
books.json.url <- getURL("https://raw.githubusercontent.com/rg563/DATA607/master/Assignments/Assignment%205/books.json")
books.json <- fromJSON(books.json.url)
class(books.json)
class(books.json$ChemEngBooks)
books.json
```

```{r json-df, eval=TRUE, warning=FALSE, comment=NA}
books.json.df <- books.json$ChemEngBooks
books.json.df
```

```{r json-datatable, eval=TRUE, warning=FALSE, comment=NA}
datatable(books.json.df)
```

## Conclusions

The major findings and differences between each file type are summarized below:

* Despite the very different file structures, HTML and JSON files were read into R using the same R code and functions (except for fromJSON and readHTMLTable functions)
* The process for scraping an XML file was more involved and a bit more challenging than the other file types. However, once the file structure of a XML file was understood, it was relatively straight forward.
* After all of the scraping was complete, all of the files produced data tables that were identical.